---
title: "Population simulations under neutral demography"
# author: Tom Ellis
date: October 2021
output:
  bookdown::pdf_book:
    toc: no
  bookdown::word_document2: null
  html_document:
    df_print: paged
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
documentclass: article
# header-includes:
# fontfamily: mathpazo
# fontsize: 12pt
# geometry: margin=1in
# bibliography: /datadisk/Dropbox/bibtex_files/tellis.bib
# csl: /datadisk/Dropbox/bibtex_files/am-nat.csl
# endnote: no
abstract: ""
editor_options:
  chunk_output_type: console
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache=T, cache.lazy=T)

library("tidyverse")
library("ggpubr")
library("simmiad")
options(warn=-1)
```

This RMarkdown document illustrates how to plot the results of simulations
of wild emmer wheat populations evolving under neutral demography only (via
limited outcrossing and seed dispersal) to recreate figures 3, S11, S12, S13 and
S14.
Simulations are generated using the R package `simmiad` v1.10 available from
[GitHub](https://github.com/ellisztamas/simmiad)
or [Zenodo](https://zenodo.org/record/4762083).

Here is a summary of input parameters for these simulations:

- 1 initial genotype
- 2000 generations
- Seed dispersal of 0.5, 0.75, 1 or 2m
- Outcrossing rates of 0.25%, 0.05%, 1% or 2%
- Densities of 1, 3 or 5 plants/m^2^
- Transects have 30 sampling points, 5 metres apart
- 100 replicate simulations for each parameter combination


```{r import-simulations}

# Import summary of the simulations from
# "simulations/005.summarise_simulations.R"
sims <- readRDS("data_analysis/simulations/output/simmiad_summary.Rds")
```

```{r format-sampling-points}
#' Format genotype and position data
#' Tom Ellis April 2021
#'
#' This script formats data for comparing with simulations and microhabitat sorting.
#' Crucially it ensures there is a row for every sampling point in every year, so
#' that there are the same number of rows for each year.

# Import data one which genotypes are found at each sampling point in each year.
years <- c(1984, 1988, 1992, 1996, 2002, 2014, 2016, 2018, 2020)

# Import data on genotype occupancy.
# Filter to give unique ID, sampling point, unique genotype, year, transect and distance
obs_geno <- read_csv(
  "data/output.csv",
  col_types = cols()
) %>%
  select(Sample, Position, IGG, Year, Habitat, dist) %>%
  # Zavitan is the ref genome control; TTD is plants from other populations.
  filter(Position != "Zavitan" & Position != "TTD")

# Ensure you have the same number of entries for each transect-year
obs_geno <- obs_geno %>%
  # Column for transect labels
  mutate(
    transect = ifelse(
      test = grepl("West", Position),
      yes = "A West",
      no = substr(Position, 1 ,1)
    )
  ) %>%
  # Join the observed data to a synthetic dataframe that has all possible years and
  # sampling points. If any are missing in reality, these will appear as NA.
  right_join(
    expand.grid(Position= unique(obs_geno$Position), Year=as.character(years)),
    by = c("Position", "Year")
  ) %>%
  # Fill in the missing data for sampling points that were empty in certain years
  group_by(Position) %>%
  fill(dist, transect, Habitat, .direction = 'downup') %>%
  ungroup() %>%
  filter(!is.na(dist)) %>%
  # reorder
  arrange(Year, transect, dist)
```

```{r decay-identity-through-space}
#' Calculate the probability of observing matching genotypes at
#' sampling points at increasing distances, averaged over years.
#' Sampling points are grouped into 5m bins

# Empty list with an element for each transect.
obs_di <- vector('list', 4)
names(obs_di) <- LETTERS[1:4]
# Loop over all four transects
for(tr in LETTERS[1:4]){
  d <- obs_geno %>%
    filter(transect==tr)
  # Dataframe giving all pairs of sampling points in the transect
  # Shows how often two sampling points 'matched' (same IGG at both point in a year)
  # This is averaged over years (n=number of years they could be compared)
  di <- distance_identities(
    genotypes = split(d$IGG, d$Year), 
    positions = split(d$dist, d$Year)[[1]]
  ) %>% 
    filter( !is.nan(matches))
  # Bin those match probabilities into 5m bins
  obs_di[[tr]] <- di %>%
    mutate(bin = cut(distances, seq(0,max(distances)+5,5))) %>% 
    group_by(bin) %>% 
    summarise(
      mean=mean(matches),
      n = n(),
      .groups='drop'
    ) %>%
    mutate(
      distance =  (1:nrow(.)) * 5,
      transect = tr
    )
}
# Join the list elements into a single dataframe
obs_di <- do.call('rbind', obs_di)
```

```{r decay-identity-through-time}
# Calculate the probability of plants of the same DGG in the same sampling 
# point at harvests separated by different number of years.

# Lists of unique sample points in each transect
sampling_pts <- lapply(LETTERS[1:4], function(x) unique(obs_geno$Position[obs_geno$transect == x]))
names(sampling_pts) <- LETTERS[1:4]
sampling_pts <- lapply(sampling_pts, na.exclude)

# Empty list to store transition data for each transect.
stability_all <- vector('list', 4)
names(stability_all) <- LETTERS[1:4]

# Loop over transects to retrieve transitions
for(t in LETTERS[1:4]){
  # empty vector to store transitions for this transect
  this_trans <- vector("list", (length(years) *(length(years)-1)) / 2) 
  # loop over the seven transitions between eight years.
  # For each pair, return the labels of each year, prob. of observing the same IGG in
  # both years, and the number of pairs that could be compared.
  counter <- 1
  for(i in 1:length(years)){
    for (j in 2:length(years)) {
      if(i < j){
        # Get genotype occupancies for year i, and index which are also in sampling_pts
        year_one <- obs_geno %>% 
          filter(transect == t & Year == years[i])
        ix <- match(sampling_pts[[t]],
                    year_one %>% select(Position) %>% pull()
        )
        # Get genotype occupancies for year j, and index which are also in sampling_pts
        year_two <- obs_geno %>% 
          filter(transect == t & Year == years[j])
        jx <- match(sampling_pts[[t]],
                    year_two %>% select(Position) %>% pull()
        )
        # Calculate stability for this transition, and the number of valid comparisons.
        this_trans[[counter]] <- c(
          year_one = years[i],
          year_two = years[j],
          stability = transect_stability(year_one$IGG[ix], year_two$IGG[jx]),
          n = (!is.na(ix) & !is.na(jx)) %>% sum()
        )
        counter <- counter +1
      }
    }
  }
  # Bind list, and add column for transect.
  # Add columns for the transect and number of years being compared
  this_trans <- do.call('rbind', this_trans) %>%
    as.data.frame() %>% 
    add_column(transect = t, .before = 'stability') %>% 
    mutate(
      n_years = year_two - year_one
    )
  # add_column(transition = years[i+1], .before = 'stability')
  # Send to bigger list
  stability_all[[t]] <- this_trans
}
# Bind list and add st error
stability_all <- do.call("rbind", stability_all)
stability_all$sterr <- sqrt(stability_all$stability * (1-stability_all$stability)) / (stability_all$n-1)

obs_stability <- stability_all %>% 
  group_by(transect) %>% 
  mutate(bins = cut(n_years, seq(1,40,4))) %>% 
  group_by(transect, n_years) %>% 
  summarise(
    stability = mean(stability),
    n = n(),
    .groups='drop'
  )
```


```{r overlap-with-obs, include=FALSE}
#' Tom Ellis, May 2021
#' Calculate how often simulations overlap with observed data for the four transects.
#' This is done for each pair of distance between sampling points and for each pair of numbers of years.
#' Simulations are compared with the *minimum* clustering/stability among transects for each pair of data,
#' so this reflects overlap with *any* observed transect.

base_folder <- "data_analysis/simulations/output/"
folders <- list.dirs(base_folder)

# Return a data.frame of parameter values for each simulation.
parameters <- vector("list", length = length(folders)-1)
for(f in 2:length(folders)){
  p <-  read.csv(paste(folders[f], "/parameters.csv", sep="")) %>%
    filter(
      parameter %in% c(
        'mean_dispersal_distance', 'outcrossing_rate', 'n_generations', 'n_starting_genotypes', 'density', 'dormancy'
      )
    )
  vals <- p$value
  names(vals) <- p$parameter
  parameters[[f]] <- vals
}
parameters <- parameters %>% 
  do.call(what = 'rbind') %>%
  as.data.frame()

# SPATIAL CLUSTERING
# Get the minimum stability observed in any transect for any pair of years.
min_obs_clustering <- obs_di %>%
  filter(distance <= 60) %>% 
  select(transect, distance, mean) %>% 
  pivot_wider(names_from = transect, values_from = mean) %>% 
  select(A:D) %>% 
  pmap(.f = min) %>% 
  unlist()
# Loop through each simulation output and count how many simulation replicates
# gave a higher probability of observing identical genotypes for each pair of 
# distances.
p_spatial <- vector("list", length = length(folders)-1 )
pb <- txtProgressBar(min = 2, max = length(folders), style = 3)   
for(f in 2:length(folders)){
  # Open the parameter file
  this_sim <- read.csv(paste(folders[f], "/distance_identity.csv", sep=""))
  this_sim <- this_sim[,1:12] %>% t() # take the first 36 years and transpose
  # How often are simulations greater than observed?
  this_comparison <- min_obs_clustering < this_sim
  p_spatial[[f]] <- rowMeans(this_comparison, na.rm = T)
  
  setTxtProgressBar(pb, f)
}
close(pb) # Close the connection
p_spatial <- p_spatial %>% 
  do.call(what = "rbind") %>% 
  as.data.frame() %>% 
  `colnames<-`(seq(5,60,5)) %>% 
  cbind(parameters) %>% 
  pivot_longer(`5`:`60`, names_to ="distance", values_to = "prob") %>% 
  mutate(
    distance = as.numeric(distance),
    dormancy = as.factor(dormancy),
    outcrossing_rate = as.factor(outcrossing_rate),
    density = as.factor(density)
  )


# TEMPORAL STABILITY
# Get the minimum stability observed in any transect for any pair of years.
min_obs_stability <- obs_stability %>%
  pivot_wider(names_from = transect, values_from = stability) %>% 
  select(A:D) %>% 
  pmap(.f = min) %>% 
  unlist()
# Loop through each simulation output and count how many simulation replicates
# gave a higher probability of observing identical genotypes for each pair of 
# years.
p_stability <- vector("list", length = length(folders)-1)
pb <- txtProgressBar(min = 2, max = length(folders), style = 3)   
for(f in 2:length(folders)){
  # Open the parameter file
  this_sim <- read.csv(paste(folders[f], "/stability.csv", sep=""))
  this_sim <- this_sim[,seq(2,36,2)] %>% t() # take the first 36 years and transpose
  # How often are simulations greater than observed?
  this_comparison <- min_obs_stability < this_sim
  p_stability[[f-1]] <- rowMeans(this_comparison, na.rm = T)
  setTxtProgressBar(pb, f)
}
close(pb) # Close the connection
# Make it a tidy table
p_stability <- p_stability %>% 
  do.call(what = "rbind") %>% 
  as.data.frame() %>% 
  `colnames<-`(seq(2,36,2)) %>% 
  cbind(parameters) %>% 
  pivot_longer(`2`:`36`, names_to ="n_years", values_to = "prob") %>% 
  mutate(
    n_years = as.numeric(n_years),
    dormancy = as.factor(dormancy),
    outcrossing_rate = as.factor(outcrossing_rate),
    density = as.factor(density)
  )
```

Figure \@ref(fig:plot-observed) shows main figure 3.

```{r plot-observed, fig.height=8/2.54, fig.width=16.9/2.54, fig.cap="Spatial clustering and temporal stability at Ammiad. Points show the probability of observing genotypes of the same DGG at (A) pairs of sampling points in the same year at increasing distances along a transect and (B) the same sampling point between pairs of years. Grey regions indicate 95% confidence intervals over populations simulated under conditions closest to empircal estimates (3 plants/m^2^, mean seed dispersal of 1m, 0.5% outcrossing per generation, 30% seeds germinating in second year after dispersal). In A, points show the average for sampling points in bins of 5m."}

sim_curve <- list(
  distance_identity =  sims$distance_identity$through_time %>% 
    filter(
      mean_dispersal_distance == 1,
      outcrossing_rate == 0.005, 
      density == 3,
      dormancy == 0.3
    ),
  stability = sims$stability$through_time %>% 
    filter(
      mean_dispersal_distance == 1,
      outcrossing_rate == 0.005,
      density == 3,
      dormancy == 0.3,
      generation <=36
    ) %>% 
    mutate(stability = mean)
)

p1 <- ggplot(obs_di, aes(x=distance, y = mean, colour=transect)) + 
  geom_line() +
  geom_point() +
  labs(
    x= "Distance (m)",
    y = "Prob. same DGG"
  ) +
  geom_ribbon(
    data=sim_curve$distance_identity, 
    aes(x=distance, ymin=lower, ymax = upper),
    colour = 'grey', fill="grey", alpha =0.5
    ) +
  # geom_line(data=sim_curve$distance_identity, aes(x=distance, y=lower), colour="grey") + 
  # geom_line(data=sim_curve$distance_identity, aes(x=distance, y=upper), colour="grey") + 
  # lims(
  #   x = c(0,200)
  # ) + 
  scale_colour_manual(values = c("#4285f4", "#ea4335", "#fbbc05", "#34a853"))
# scale_color_brewer(palette="Dark2")

p2 <- ggplot(obs_stability, aes(x=n_years, y = stability, colour=transect)) + 
  geom_line() +
  geom_point() +
  labs(
    x= "Number of years",
    y = "Prob. same DGG"
  ) +
  geom_ribbon(
    data=sim_curve$stability, 
    aes(x=generation, ymin=lower, ymax = upper),
    colour = 'grey', fill="grey", alpha = 0.5
    ) +
  # geom_line(data=sim_curve$stability, aes(x=generation, y=lower), colour="grey") + 
  # geom_line(data=sim_curve$stability, aes(x=generation, y=upper), colour="grey") + 
  scale_colour_manual(values = c("#4285f4", "#ea4335", "#fbbc05", "#34a853"))
# scale_color_brewer(palette="Dark2")

ggarrange(p1, p2,
          common.legend = TRUE, labels = "AUTO"
)
```

Figures \@ref(fig:spatial-sims), \@ref(fig:spatial-overlap), \@ref(fig:stability-sims) and \@ref(fig:temporal-overlap) correspond to figures S11, S12, S13 and S14.

```{r spatial-sims, fig.cap="Spatial clustering along a transect in simulated populations. Subplots show the probability of observing plants of the same DGG at pairs of sampling points at increasing distances, averaged over 100 replicate simulations. Panels indicate increasing mean seed dispersal distances (left to right) and increasing outcrossing rates (top to bottom). 'Dormancy' indicates the proportion of seeds drawn from the seed bank in each generation, and 'density' the number of plants per square metre."}
# New facet label names for dispersal distances
disp.labs <- paste(c(0.5, 0.75, 1, 2),"m",sep="")
names(disp.labs) <- c(0.5, 0.75, 1, 2)
# New facet label names for outcrossing
out.labs <- paste(c(0.0025, 0.005, 0.01, 0.02)*100, "%", sep="")
names(out.labs) <- c(0.0025, 0.005, 0.01, 0.02)

sims$distance_identity$through_time %>% 
  mutate(
    density = as.factor(density),
    dormancy = as.factor(dormancy)
  ) %>% 
  ggplot(aes(x = distance, y = mean, colour = density, linetype = dormancy)) + 
  geom_line() + 
  # geom_line(data = obs_minima$spatial, mapping = aes(x = distance, y = min, colour='gray'))
  lims(
    x=c(0,50)
  )+
  labs(
    x="Distance between sampling points (m)",
    y = "Prob. same DGG",
    linetype="Dormancy",
    color ="Density"
  )+
  theme_bw() + 
  scale_color_brewer(palette="Dark2") +
  facet_grid(
    outcrossing_rate ~ mean_dispersal_distance,
    labeller = labeller(outcrossing_rate = out.labs, mean_dispersal_distance = disp.labs)
  )
```

```{r spatial-overlap, fig.width=16.9/2.54, fig.cap="Overlap in spatial clustering between observed data and populations simulated under neutral demography. Plots show the proportion of replicate simulations where pairs sampling points at different distances were more likely to be occupied by plants of the same DGG than in any observed transect. Panels indicate increasing mean seed dispersal distances (left to right) and increasing outcrossing rates (top to bottom). 'Dormancy' indicates the proportion of seeds drawn from the seed bank in each generation, and 'density' the number of plants per square metre."}
p_spatial %>% 
  ggplot(aes(x = distance, y = prob,
             colour=density, linetype=dormancy
             # group=paste(outcrossing_rate,dormancy)
  )
  ) +
  geom_line() +
  geom_point() + 
  theme_bw()+ 
  labs(
    x = "Distance between sampling points",
    y = "N. replicates overlapping with observed data",
    linetype="Dormancy",
    color ="Density"
  ) +
  # scale_color_brewer(palette="Set1") +
  scale_color_brewer(palette="Dark2") +
  facet_grid(
    outcrossing_rate ~ mean_dispersal_distance,
    labeller = labeller(outcrossing_rate = out.labs, mean_dispersal_distance = disp.labs)
  )
```

```{r stability-sims, fig.cap="Temporal stability in simulated populations. Subplots show the probability of observing plants of the same DGG at a single sampling in two different years, averaged over 100 replicate simulations. Panels indicate increasing mean seed dispersal distances (left to right) and increasing outcrossing rates (top to bottom). 'Dormancy' indicates the proportion of seeds drawn from the seed bank in each generation, and 'density' the number of plants per square metre."}
sims$stability$through_time %>% 
  mutate(
    density = as.factor(density),
    dormancy = as.factor(dormancy)
  ) %>% 
  ggplot(aes(x = generation, y = mean, colour = density, linetype = dormancy)) + 
  geom_line() + 
  lims(
    x=c(0,40)
  )+
  labs(
    x="Years between harvests (m)",
    y = "Prob. same DGG",
    linetype="Dormancy",
    color ="Density"
  )+
  theme_bw() + 
  scale_color_brewer(palette="Dark2") +
  facet_grid(
    outcrossing_rate ~ mean_dispersal_distance,
    labeller = labeller(outcrossing_rate = out.labs, mean_dispersal_distance = disp.labs)
  )
```

```{r temporal-overlap, fig.width=16.9/2.54, fig.cap="Probabilities that observed temporal stability could be generated by neutral demography. Plots show the proportion of replicate simulations where a single sampling point sampled in different harvest years was more likely to be occupied by plants of the same DGG than in any observed transect. Panels indicate increasing mean seed dispersal distances (left to right) and increasing outcrossing rates (top to bottom). 'Dormancy' indicates the proportion of seeds drawn from the seed bank in each generation, and 'density' the number of plants per square metre."}
p_stability %>% 
  ggplot(aes(
    x = n_years, y = prob,
    colour=density, linetype=dormancy
    # group=paste(outcrossing_rate,dormancy)
  )
  ) + 
  geom_line() +
  theme_bw()+ 
  labs(
    x = "Years between harvests",
    y = "N. replicates overlapping with observed data",
    linetype="Dormancy",
    color ="Density"
  ) +
  # scale_x_continuous(breaks = waiver()) +
  scale_color_brewer(palette="Dark2") +
  # scale_color_brewer(palette="Set1") + 
  facet_grid(
    outcrossing_rate ~ mean_dispersal_distance,
    labeller = labeller(outcrossing_rate = out.labs, mean_dispersal_distance = disp.labs)
  )
```
