---
title: "Are identical genotypes found in the same microhabitats?"
author: "Tom Ellis"
date: "17/03/2021"
output:
  bookdown::pdf_book:
    toc: no
  html_document: default
---

We would like to know whether plants of the same genotype tend to be found in the same microhabitat, which would indicate selection for microhabitat. To do this we calculate how many pairs of plants of matching genotypes are found within the same microhabitat. We compare this observed value to values from some kind of null distribution.


```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache=T, cache.lazy=T, fig.align='center', warning = F)

library("tidyverse")

source("data_analysis/habitat_permutations/permutation_functions.R")

source("data_processing/import_field_occupancy_data.R")

```

One way to do this is to:

1. Line all the habitat labels up across the transects, and join the ends to make a circle.
2. Randomly permute the order of the habitats, but keeping matching labels together.
3. Randomly rotate the vector some distance (so that the permuted habitat labels do not always coincide with the start of the transect).

Results are shown in figure \@ref(fig:plot-permutations), corresponding to
figure S8 in the text.

```{r calculate-permutations}
permutations <- vector('list', length(years))

# Calculate sorting for each year.
# Returns a list of data.frame from `habitat_rotations()` with an entry for each
# year.
for(y in 1:length(years)){
  # Pull out data for a single year.
  this_year <- obs_geno %>% 
    filter(Year == years[y], transect != "C", !is.na(Habitat)) %>% 
    arrange(transect, dist)
  # Perform rotations and calculated sorting by microhabitat.
  permutations[[y]] <- habitat_permutations(
    genotype = this_year$DGG,
    habitat = this_year$Habitat,
    nreps=1000
  )
}
# Add columns for year and concatenate
permutations <- lapply(1:length(years), function(y) cbind(year=years[y], permutations[[y]]))
names(permutations) <- as.character(years)
```

```{r plot-permutations, fig.cap="Sorting of identical genotypes by microhabitat comparing observed values (red) to values obtained by permuting habitat labels (violins)."}

pvals <- sapply(permutations, function(x){
  n <- nrow(x) -1
  mean(x$matches[n + 1] < x$matches[1:n])
})

do.call('rbind', permutations) %>% 
  filter(type == "permuted") %>% 
  
  ggplot(aes(x=factor(year), y = matches)) + 
  geom_violin(adjust=1.5) +
  geom_point(
    data = do.call('rbind', permutations) %>% filter(type == "observed"),
    aes(x=factor(year), y = matches),
    colour = "red") + 
  scale_color_brewer(palette="Dark2") + 
  labs(
    x="Year",
    y=expression("F"[st])
  ) +
  theme_bw()

ggsave(
  "data_analysis/habitat_permutations/habitat_permutations.pdf",
  device = "pdf",
  units = "cm", height  = 12, width = 12
)

```

